---
title: "Severe Weather Costs 1950 - 2011"
author: "Jeff Anderson"
date: "7/24/2018"
output: html_document
---

# SEVERE WEATHER COSTS FROM 1950 TO 2011  

The main purpose of this analysis is to total up the human and monetary costs of severe weather from 1950 to 2011.  This study will summarize and rank the following measures per event in descending order (most expensive firsts)  The following costs are summarized:

- Total Number Injuries (worst 10)
- Total Number of Fatalities (worst 10)
- The Sum of Crop and Property Damage in Billions (worst 10)

## DATA

The source of this data is the NOAA Storm Database

- Documentation about the databsae can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)  
- FAQ about the database can be found [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)
- The actual data used for this study is [here](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

## Load Data

If the data does not exist in the folder download and unzip it.  Then, regardless load all the data  

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(R.utils)

if (!file.exists("StormData.csv")) {
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2","StormData.csv.bz2")
  bunzip2("StormData.csv.bz2")
}

raw.data = read.table("StormData.csv", sep=",", header = T, stringsAsFactors = F)

# list column headings  
print(colnames(raw.data))
# get table structure
print(str(raw.data))
```
## Data Analysis / Transformations

Based on the list of columns I only have interest in date, event, fatality, injury and cost related items where there are amounts > 0

```{r}
base.data = raw.data %>% 
    select(BGN_DATE,EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP) %>%
    filter(FATALITIES>0 | INJURIES>0 | PROPDMG>0, CROPDMG>0)
```

Now we need to see the codes used for costs

```{r}
print(unique(base.data$PROPDMGEXP))
print(unique(base.data$CROPDMGEXP))
```

Looks like we have some mixed case values, let us fix that:

```{r}
base.data$PROPDMGEXP = toupper(base.data$PROPDMGEXP)
base.data$CROPDMGEXP = toupper(base.data$CROPDMGEXP)
```

Now let's see that there are no lower case values..   

```{r}
print(unique(base.data$PROPDMGEXP))
print(unique(base.data$CROPDMGEXP))
```

Now we see some valid values and some invalid values for cost codes:  
- B = billions - factor, will be 1  
- M = millions - factor, will be .001  
- T = thousands - factor, will be .000001  
- other values - unknown, factor will be 0  

Based on the above let's standardize on Billions.  This will require a factor to multiply the costs by to convert all costs to Billions (factors above).  From there we calculate total costs in billions by multiplying crop and property costs by each factor and adding the two products to get total cost.

```{r}
base.data = base.data %>%
  mutate(prop.factor = ifelse(PROPDMGEXP == "B", 1, 
                         ifelse(PROPDMGEXP=="M", .001,
                            ifelse(PROPDMGEXP=="K", .000001,0)))) %>%
  mutate(crop.factor = ifelse(CROPDMGEXP == "B", 1, 
                          ifelse(CROPDMGEXP=="M", .001,
                            ifelse(CROPDMGEXP=="K", .000001,0)))) %>%
  mutate(total.cost = (PROPDMG * prop.factor) + (CROPDMG * crop.factor))

print(head(base.data))
```

Now let us look at the different weather events  

```{r}
print(unique(base.data$EVTYPE))
```

Based on this list it looks like several weather events could be combined 

```{r}
# organaize hurricanes
cur.type = grepl("hurricane", base.data$EVTYPE, ignore.case = T) 
base.data[cur.type,2] = "HURRICANE"
# organaize thunderstorm
cur.type = grepl("thunderstorm", base.data$EVTYPE, ignore.case = T) 
base.data[cur.type,2] = "THUNDERSTORM"
# organaize flooding
cur.type = grepl("flood", base.data$EVTYPE, ignore.case = T) 
base.data[cur.type,2] = "FLOOD"
# organaize flooding
cur.type = grepl("tornado", base.data$EVTYPE, ignore.case = T) 
base.data[cur.type,2] = "TORNADO" 
#organaize heat
cur.type = grepl("heat", base.data$EVTYPE, ignore.case = T) 
base.data[cur.type,2] = "HEAT"
#combine winters storms/blizzards
cur.type = grepl("blizz", base.data$EVTYPE, ignore.case = T) 
base.data[cur.type,2] = "WINTER STORM"
cur.type = grepl("winter storm", base.data$EVTYPE, ignore.case = T) 
base.data[cur.type,2] = "WINTER STORM"
# assuming tstm = tropical storm
cur.type = grepl("tstm win", base.data$EVTYPE, ignore.case = T) 
base.data[cur.type,2] = "TROPICAL STORM"
```

Now let us look at the updated list of events  

```{r}
print(unique(base.data$EVTYPE))
```

## Summarize Final Dataset and Subsets

```{r}
sum.data = base.data %>%
  select(EVTYPE,FATALITIES,INJURIES,total.cost) %>%
  group_by(EVTYPE) %>%
  summarise(TOTAL_FATALITIES = sum(FATALITIES),
            TOTAL_INJURIES = sum(INJURIES),
            TOTAL_COST_BILLIONS = sum(total.cost))

print(head(sum.data))

sum.fatalities = sum.data %>% 
  select(EVTYPE, TOTAL_FATALITIES) %>%
  arrange(desc(TOTAL_FATALITIES), EVTYPE)

print(head(sum.fatalities))

sum.injuries = sum.data %>% 
  select(EVTYPE, TOTAL_INJURIES) %>%
  arrange(desc(TOTAL_INJURIES), EVTYPE)

print(head(sum.injuries))

sum.cost = sum.data %>% 
  select(EVTYPE, TOTAL_COST_BILLIONS) %>%
  arrange(desc(TOTAL_COST_BILLIONS), EVTYPE)

print(head(sum.cost))
```

## Summary Graphs

```{r}
# Total Fatalities Graph
tf = ggplot(data=sum.fatalities[1:10,], 
       aes(x=reorder(EVTYPE, -TOTAL_FATALITIES), y=TOTAL_FATALITIES))+
  geom_col(fill="darkred")+
  theme(axis.text.x=element_text(angle = -90, hjust = 0))+
  labs(title="Fatalities by Weather Event", x="Weather Event", y="Number of Fatalities")

print(tf)

# Total Injuries Graph
ti = ggplot(data=sum.injuries[1:10,], 
       aes(x=reorder(EVTYPE, -TOTAL_INJURIES), y=TOTAL_INJURIES))+
  geom_col(fill="darkblue")+
  theme(axis.text.x=element_text(angle = -90, hjust = 0))+
  labs(title="Injuries by Weather Event", x="Weather Event", y="Number of Injuries")

print(ti)

# Total Cost Graph
tc = ggplot(data=sum.cost[1:10,], 
       aes(x=reorder(EVTYPE, -TOTAL_COST_BILLIONS), y=TOTAL_COST_BILLIONS))+
  geom_col(fill="darkgreen")+
  theme(axis.text.x=element_text(angle = -90, hjust = 0))+
  labs(title="Costs (Billions) by Weather Event", 
       x="Weather Event", 
       y="Property/Crop Cost (Billions)")

print(tc)
```

# Final Summary  

Floods are the most costly weather event in all respects.  Huricanes, tornadoes, heat and ice storms are all very costly as well. 

The RPub document is [here](http://rpubs.com/andersjh/repo_weather_data)
